---
title: "Unit8CaseStudy"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ShowInstalledPackages1, include=FALSE, echo=FALSE, eval=TRUE}
installed.packages() 
sessionInfo()
```

First we experiment with accessing the 2012 men's results. The first ten lines are viewed and we can see that we have successfully imported the data.

```{r Explore}
library(rvest)
library(magrittr)

ubase = "http://www.cherryblossom.org/"
url = paste(ubase, "results/2012/2012cucb10m-m.htm", sep = "")
doc = read_html(url)

els <- doc %>%
  html_node('pre') %>%
  html_text() %>%
  strsplit("\\r\\n")
els <- els[[1]]
els[1:10]

```

```{r Explore}
url = paste(ubase, "results/1999/cb99m.html", sep = "")
doc = read_html(url)

els <- doc %>%
  html_node('pre') %>%
  html_text() 
  
els <- strsplit(els, '\\n')[[1]]
els[1:10]

```
```{r Explore}
url = paste(ubase, "results/2000/Cb003m.htm", sep = "")
doc = read_html(url)

 els <- doc %>%
   html_node('body') %>%
   html_text() 
#   
els <- strsplit(els, "\r\n")[[1]]
els[1:10]

```

Next we attempted to create a function to scrape data for all 14 years we are interested in assessing. The function is shown below.

```{r ExtractResTable1}
extractResTable1 =
       # Retrieve data from web site, find preformatted text,
       # return as a character vector.
function(url)
{
  doc = read_html(url)
  els <- doc %>%
    html_node('pre') %>%
    html_text() 
   
  els <-  strsplit(els, "\\r\\n")[[1]]
  return(els)
}
```

Let's test our extraction function on the 2012 data. We see that it works for this year.

```{r, include=FALSE, echo=FALSE}
men2012 <- extractResTable1("http://cherryblossom.org/results/2012/2012cucb10m-m.htm")
men2012[1:10]
```

The function works! Next we created a list of all of the URls for the male racers.

```{r WebScrapeMen, include=FALSE, echo=FALSE}
ubase = "http://www.cherryblossom.org/"
menURLs = c(paste(ubase, "results/", 1999, "/", "cb99m.html", sep = ""),
            paste(ubase, "results/", 2000, "/", "Cb003m.htm", sep = ""),
            paste(ubase, "results/", 2001, "/", "oof_m.html", sep = ""),
            paste(ubase, "results/", 2002, "/", "oofm.htm", sep = ""),
            paste(ubase, "results/", 2003, "/", "CB03-M.HTM", sep = ""),
            paste(ubase, "results/", 2004, "/", "men.htm", sep = ""),
            paste(ubase, "results/", 2005, "/", "CB05-M.htm", sep = ""),
            paste(ubase, "results/", 2006, "/", "men.htm", sep = ""),
            paste(ubase, "results/", 2007, "/", "men.htm", sep = ""),
            paste(ubase, "results/", 2008, "/", "men.htm", sep = ""),
            paste(ubase, "results/", 2009, "/", "09cucb-M.htm", sep = ""),
            paste(ubase, "results/", 2010, "/", "2010cucb10m-m.htm", sep = ""),
            paste(ubase, "results/", 2011, "/", "2011cucb10m-m.htm", sep = ""),
            paste(ubase, "results/", 2012, "/", "2012cucb10m-m.htm", sep = ""))
```

Likewise, a list was created for the women.

```{r WebScrapeWomen, include=FALSE, echo=FALSE}
ubase = "http://www.cherryblossom.org/"
womenURLs = c(paste(ubase, "results/", 1999, "/", "cb99f.htm", sep = ""),
            paste(ubase, "results/", 2000, "/", "Cb003f.htm", sep = ""),
            paste(ubase, "results/", 2001, "/", "oof_f.html", sep = ""),
            paste(ubase, "results/", 2002, "/", "ooff.htm", sep = ""),
            paste(ubase, "results/", 2003, "/", "CB03-F.HTM", sep = ""),
            paste(ubase, "results/", 2004, "/", "women.htm", sep = ""),
            paste(ubase, "results/", 2005, "/", "CB05-F.htm", sep = ""),
            paste(ubase, "results/", 2006, "/", "women.htm", sep = ""),
            paste(ubase, "results/", 2007, "/", "women.htm", sep = ""),
            paste(ubase, "results/", 2008, "/", "women.htm", sep = ""),
            paste(ubase, "results/", 2009, "/", "09cucb-F.htm", sep = ""),
            paste(ubase, "results/", 2010, "/", "2010cucb10m-f.htm", sep = ""),
            paste(ubase, "results/", 2011, "/", "2011cucb10m-f.htm", sep = ""),
            paste(ubase, "results/", 2012, "/", "2012cucb10m-f.htm", sep = ""))
```

Next we created tables for every year, we looked at the lengths for the men tables.  The years 1999, 2000, and 1999 have lengths of one. The extractResTable function didn't work properly for all years. We'll need another approach for those that didn't pull in correctly.

```{r}
menTables = lapply(menURLs, extractResTable1)
names(menTables) = 1999:2012
sapply(menTables, length)
```

A new function is created with a different methodology for the year 1999 The new function is applied and we again assess the number of records in each dataset. The year 1999 appears to have been correctly pulled in this time.

```{r ExtractResTable2}
extractResTable2 =
  # Retrieve data from web site, 
  # find the preformatted text,
  # and return as a character vector.
function(url, year = 1999)
{
  doc = read_html(url)
  els <- doc %>%
    html_node('pre') %>%
    html_text()
  
  if (year == 1999) {
    els <- strsplit(els, "\\n")[[1]]
    }
  else {
    els = strsplit(els, "\\r\\n")[[1]]
    }

  return(els)
}

years = 1999:2012
menTables = mapply(extractResTable2, url = menURLs, year = years)
names(menTables) = years
sapply(menTables, length)
```

The function is further refined to correctly pull the 2000 data.

```{r ExtractResTable3}
extractResTable3 =
  #
  # Retrieve data from web site, 
  # find the preformatted text,
  # and write lines or return as a character vector.
  #
  function(url = "http://www.cherryblossom.org/results/2009/09cucb-F.htm", year = 1999, sex = "male", file = NULL)
  {
    doc = htmlParse(url)

    if (year == 2000) {
      # Get preformatted text from 4th font element
      # The top file is ill formed so the <pre> search doesn't work.
      ff = getNodeSet(doc, "//font")
      txt = xmlValue(ff[[4]])
      els = strsplit(txt, "\r\n")[[1]]
    }
    else if (year == 2009 & sex == "male") {
      # Get preformatted text from <div class="Section1"> element
      # Each line of results is in a <pre> element
      div1 = getNodeSet(doc, "//div[@class='Section1']")
      pres = getNodeSet(div1[[1]], "//pre")
      els = sapply(pres, xmlValue)
    }
    else {
      # Get preformatted text from <pre> elements
      pres = getNodeSet(doc, "//pre")
      txt = xmlValue(pres[[1]])
      els = strsplit(txt, "\r\n")[[1]]   
    } 
    
    if (is.null(file)) return(els)
    # Write the lines as a text file.
    writeLines(els, con = file)
  }

years = 1999:2012
menTables = mapply(extractResTable3, url = menURLs, year = years)
names(menTables) = years
sapply(menTables, length)
```

Let's try the new function on the women's data.

```{r}
years = 1999:2012
womenTables = mapply(extractResTable2, url = womenURLs, year = years)
names(womenTables) = years
sapply(womenTables, length)
```

After some searching on the web, the following approach was found at the github repo list below. We test to see if it still works.

https://github.com/bthomasbailey/cherry-blossom-run/blob/master/scrapeWeb.R

```{r extractSingleItem}
extractSingleItem <- function(strLine, pattern){
    itemVal <- str_match(strLine, pattern)[1, 1]
    len <- nchar(itemVal)
    
    if (!is.na(itemVal)) {
        loc <- regexpr(itemVal, strLine)
        newStrLine <- substr(strLine, start = loc + len, stop = nchar(strLine))   
    } else {
        newStrLine <- strLine
    }
    
    return (c(itemVal, newStrLine))
}
```

```{r parseLine}
parseLine <- function(strLine){
    
    placeOut <- extractSingleItem(strLine, "\\d+")
    place <- str_trim(placeOut[1])
    newStrLine <- placeOut[2]
    
    divTotOut <- extractSingleItem(newStrLine, "\\d+/\\d+")
    divTot <- str_trim(divTotOut[1])
    newStrLine <- divTotOut[2]
    
    numOut <- extractSingleItem(newStrLine, "\\d+")
    num <- str_trim(numOut[1])
    newStrLine <- numOut[2]
    
    nameOut <- extractSingleItem(newStrLine, "[A-Za-z\\-\\'\\.\\s]+")
    name <- str_trim(nameOut[1])
    newStrLine <- nameOut[2]
    
    agOut <- extractSingleItem(newStrLine, "\\d+")
    ag <- str_trim(agOut[1])
    newStrLine <- agOut[2]
    
    homeTownOut <- extractSingleItem(newStrLine, "([A-Za-z]+[\\s]*)+")
    homeTown <- str_trim(homeTownOut[1])
    newStrLine <- homeTownOut[2]
    
    times <- str_match_all(newStrLine, "\\d+[\\:]{1}\\d+([\\:]{1}\\d+)?[#|*]*")[[1]]
    gunTime <- str_trim(times[1, 1])
    netTime <- str_trim(times[2, 1])
    pace <- str_trim(times[3, 1])
    
    return (c(place, divTot, num, name, ag, homeTown, gunTime, netTime, pace))
}
```

```{r extractResTable4}
extractResTable4 <- function(url, year = 1999, men = T, file = NULL) {
    #Retrieve data from web site, find preformatted text, return as chr vector
    
    require(XML)
    require(gdata)
    require(stringr)
    
    doc <- htmlParse(url)
    
    if (year == 2000) {
        #Get text from 4th font element
        #File is ill-formed so <pre> search doesn't work.
        ff <- getNodeSet(doc, "//font")
        txt <- xmlValue(ff[[4]])
        els <- strsplit(txt, "\r\n")[[1]]
    } else if (year == 2009 & men == T) {
        #The html for this is nooooo fun
        dp <- getNodeSet(doc, "//div//pre")
        nodeVals <- lapply(dp, xmlValue)
        
        eqIndex <- grep("^===", nodeVals)
        
        spacerRow <- nodeVals[eqIndex][[1]]
        vecSpacerRow <- str_split(str_trim(spacerRow), "[\\s|Â]+")[[1]]
        
        #Need to adjust Gun Time and Net Time columns b/c they are too short for some of their values
        vecSpacerRow[7] <- paste(vecSpacerRow[7],"=", sep = "")
        vecSpacerRow[8] <- paste(vecSpacerRow[8],"=", sep = "")
        
        #manually type in headers; I know this isn't ideal, but code is only useful for Men's 2009 results anyway
        vecHeaderRow <- c("Place", "Div/Tot", "Num", "Name", "Ag", "Hometown", "Gun Tim", "Net Tim", "Pace")
        
        bodyNodeVals <- nodeVals[(eqIndex + 1):(length(nodeVals) - 2)]
        
        colLens <- sapply(vecSpacerRow, nchar, USE.NAMES = F)
        bodyMat <- t(sapply(bodyNodeVals, parseLine))
        
        preHeadNodeVals <- nodeVals[(1):(eqIndex - 2)]
        vecPreHeadText <- sapply(preHeadNodeVals, gsub, pattern = "Â", replacement = " ")
        vecPreHeadText <- sapply(vecPreHeadText, str_trim)
        
        footerNodeVals <- nodeVals[(length(nodeVals) - 1):(length(nodeVals))]
        vecFooterText <- sapply(footerNodeVals, gsub, pattern = "Â", replacement = " ")
        #footerMat <- t(sapply(vecFooterText, function(x) c(x, rep(" ", 8)), USE.NAMES = F))
        
        dataMat <- rbind(vecHeaderRow, vecSpacerRow, bodyMat, deparse.level = 0)
        
        #write to Txt file
        lapply(vecPreHeadText, write, "temp2009.txt", append = T)
        write.fwf(dataMat, file = "temp2009.txt", width = colLens, colnames = F, append = T)
        lapply(vecFooterText, write, "temp2009.txt", append = T)
        
        #convert contents of txt file to chr vector
        els <- readLines("temp2009.txt")
        file.remove("temp2009.txt")
    }
    else {
        preNode <- getNodeSet(doc, "//pre")
        txt <- xmlValue(preNode[[1]])
        els <- strsplit(txt, "\r\n")[[1]]
    }
    
    if (men == T) {
        subDir <- "data/MenTxt"
    } else {
        subDir <- "data/WomenTxt"
    }
    
    if (!(is.null(file))) {
        if(!(dir.exists(subDir))) {dir.create(subDir)} 
        writeLines(els, file.path(subDir, file))
    }
    
    return(els)
}

years = 1999:2012
menTables = mapply(extractResTable4, url = menURLs, year = years, men = T, file = paste(years,".txt", sep=""))
names(menTables) = years
sapply(menTables, length)
```

```{r}
years = 1999:2012
womenTables = mapply(extractResTable4, url = womenURLs, year = years, men = F, file = paste(years,".txt", sep=""))
names(womenTables) = years
sapply(womenTables, length)
```

Now we save all of the data in .rda files.

```{r SaveData}
save(menTables, file = "data/CBMenTextTables.rda")
#save(womenTables, file = "data/CBWomenTextTables.rda")
```

Next we create some helper Functions to read back in the text files we saved previously. This is a work around to get the 1999 data to not read in as one long line.

```{r HelperFunctions}
findColLocs <- function(spacerRow) {
    
    spaceLocs <- gregexpr(" ", spacerRow)[[1]]
    rowLength <- nchar(spacerRow)
    
    if (substring(spacerRow, rowLength, rowLength) != " ") {
        return (c(0, spaceLocs, rowLength + 1))
    } else {
        return (c(0, spaceLocs))
    }
}

selectCols <- function(colNames, headerRow, searchLocs) {
    sapply(colNames,
           function(name, headerRow, searchLocs) {
               startPos <- regexpr(name, headerRow)[[1]]
               if (startPos == -1) {
                   return(c(NA, NA))
               }
               
               index <- sum(startPos >= searchLocs)
               c(searchLocs[index] + 1, searchLocs[index + 1])
           },
           headerRow = headerRow, searchLocs = searchLocs)
    
}

extractVariables <- function(file, varNames = c("name", "home", "ag", "gun", "net", "time"),
                             sex, year) {
    
    #Find the index of the footer row
    footIndex <- grep("^[[:blank:]]*[#|*]", file)
    
    #Find the index of rows that are completely blank
    blankIndex <- grep("^[[:blank:]]*$", file)
    
    if(sex == "W" & year == 2001){
        #women's file for 2001 does not contain spacer or header rows
        body <- file[-c(footIndex, blankIndex)]
        locCols<-matrix(c(13, 34, 38, 56, 35, 37, 65, 72, 57, 64, NA, NA), nrow = 2)
        colnames(locCols) <- varNames
        
    } else {
        #Find the index of the row with equal signs
        eqIndex <- grep("^===", file)    
        
        #Extract the two key rows and the data (fix men 2006 spacer row)
        spacerRow <- file[eqIndex]
        headerRow <- tolower(file[eqIndex - 1])
        
        if (year == 2006){
            locNetTime <- regexpr("net", headerRow)
            spacerRow <- paste(substr(spacerRow, 1, locNetTime - 2), 
                               substr(spacerRow, locNetTime, nchar(spacerRow)), "")
        }
        
        
        body <- file[-c(1:eqIndex, footIndex, blankIndex)]
        
        #Obtain the starting and ending positions of variables
        searchLocs <- findColLocs(spacerRow)
        locCols <- selectCols(varNames, headerRow, searchLocs)
    }
    
    Values <- mapply(substr, list(body), start = locCols[1,], stop = locCols[2,])
    colnames(Values) <- varNames
    
    invisible(Values)
}
```

```{r}
setwd("/Users/andrewabbott/Documents/Courses/MSDS7333 Quantifying the World/Unit 7-8/data")

mfilenames <- paste("MenTxt/", 1999:2012, ".txt", sep="")
menFiles <- lapply(mfilenames, readLines)
names(menFiles) <- 1999:2012

#wfilenames <- paste("WomenTxt/", 1999:2012, ".txt", sep="")
#womenFiles <- lapply(wfilenames, readLines)
#names(womenFiles) <- 1999:2012
```

```{r}
menResMat <- mapply(extractVariables, menFiles, sex = "M", year = 1999:2012)
#womenResMat <- mapply(extractVariables, womenFiles, sex = "W", year = 1999:2012)

sapply(menResMat, nrow)
#sapply(womenResMat, nrow)
```

Finally our datasets are set up correctly. The 1999 data no longer reads as one string and all of our column names are uniform. We can now proceed with analyzing the data.

```{r}
menResMat[12]
```

```{r}
setwd("~/EMD/CaseStudies/Runners/WebPages")

m2012 = read.table(file="MenTxt/2012.txt", skip = 8)

els = readLines("MenTxt/2012.txt")

els[1:10]

els2011 = readLines("MenTxt/2011.txt")
els2011[1:10]

eqIndex = grep("^===", els)
eqIndex

first3 = substr(els, 1, 3)
which(first3 == "===")

spacerRow = els[eqIndex]
headerRow = els[eqIndex - 1]
body = els[ -(1:eqIndex) ]

headerRow = tolower(headerRow)

ageStart = regexpr("ag", headerRow)
ageStart

age = substr(body, start = ageStart, stop = ageStart + 1)
head(age)

summary(as.numeric(age))
```


```{r}
blankLocs = gregexpr(" ", spacerRow)
blankLocs

searchLocs = c(0, blankLocs[[1]])

Values = mapply(substr, list(body), 
                start = searchLocs[ -length(searchLocs)] + 1, 
                stop = searchLocs[ -1 ] - 1)
```

```{r FindColLocs}
findColLocs = function(spacerRow) {

  spaceLocs = gregexpr(" ", spacerRow)[[1]]
  rowLength = nchar(spacerRow)

  if (substring(spacerRow, rowLength, rowLength) != " ")
    return( c(0, spaceLocs, rowLength + 1))
  else return(c(0, spaceLocs))
}
```

```{r SelectCols}
selectCols = 
function(colNames, headerRow, searchLocs) 
{
  sapply(colNames, 
         function(name, headerRow, searchLocs)
         {
           startPos = regexpr(name, headerRow)[[1]]
           if (startPos == -1) 
             return( c(NA, NA) )
    
           index = sum(startPos >= searchLocs)
           c(searchLocs[index] + 1, searchLocs[index + 1] - 1)
         },
         headerRow = headerRow, searchLocs = searchLocs )
}
```

```{r}
searchLocs = findColLocs(spacerRow)
ageLoc = selectCols("ag", headerRow, searchLocs) 
ages = mapply(substr, list(body), 
              start = ageLoc[1,], stop = ageLoc[2, ])

summary(as.numeric(ages))

shortColNames = c("name", "home", "ag", "gun", "net", "time")

locCols = selectCols(shortColNames, headerRow, searchLocs)

Values = mapply(substr, list(body), start = locCols[1, ], 
                stop = locCols[2, ])

class(Values)
```

colnames(Values) = shortColNames
head(Values)

tail(Values)[ , 1:3]

extractVariables = 
  function(file, varNames =c("name", "home", "ag", "gun",
                             "net", "time"))
{
       # Find the index of the row with =s
  eqIndex = grep("^===", file)
       # Extract the two key rows and the data
  spacerRow = file[eqIndex] 
  headerRow = tolower(file[ eqIndex - 1 ])
  body = file[ -(1 : eqIndex) ]
       
       # Obtain the starting and ending positions of variables
  searchLocs = findColLocs(spacerRow)
  locCols = selectCols(varNames, headerRow, searchLocs)

  Values = mapply(substr, list(body), start = locCols[1, ], 
                  stop = locCols[2, ])
  colnames(Values) = varNames
  
  invisible(Values)
}

mfilenames = paste("MenTxt/", 1999:2012, ".txt", sep = "")
menFiles = lapply(mfilenames, readLines)
names(menFiles) = 1999:2012

menResMat = lapply(menFiles, extractVariables)
length(menResMat)

sapply(menResMat, nrow)

### The 2001 results for women are missing the === and the column names.
### Can we pick it up from the 2001 men? YES! Make an exercise
#wfilenames = paste("WomenTxt/", 1999:2012, ".txt", sep = "")
#womenTables = lapply(wfilenames, readLines)

#womenTables[[3]][1:5]

#names(womenTables) = 1999:2012
#womenResMat = lapply(womenTables, extractVariables)
#head(womenResMat[[3]], 10)
#tail(womenResMat[[3]], 10)

age = as.numeric(menResMat[['2012']][ , 'ag'])

tail(age)

age = sapply(menResMat,
             function(x) as.numeric(x[ , 'ag']))


pdf("CB_BoxplotAgeByYr.pdf", width = 8, height = 5)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

boxplot(age, ylab = "Age", xlab = "Year")

par(oldPar)
dev.off()

head(menFiles[['2003']])

menFiles[['2006']][2200:2205]

selectCols = function(shortColNames, headerRow, searchLocs) {
  sapply(shortColNames, function(shortName, headerRow, searchLocs){
    startPos = regexpr(shortName, headerRow)[[1]]
    if (startPos == -1) return( c(NA, NA) )
    index = sum(startPos >= searchLocs)
    c(searchLocs[index] + 1, searchLocs[index + 1])
  }, headerRow = headerRow, searchLocs = searchLocs )
}

menResMat = lapply(menFiles, extractVariables)
#womenResMat = lapply(womenFiles, extractVariables)

age = sapply(menResMat, 
             function(x) as.numeric(x[ , 'ag']))

pdf("CB_BoxplotAgeByYrRevised.pdf", width = 8, height = 5)
oldPar = par(mar = c(4.1, 4.1, 1, 1))
boxplot(age, ylab = "Age", xlab = "Year")
par(oldPar)
dev.off()

sapply(age,  function(x) sum(is.na(x)))

age2001 = age[["2001"]]

grep("^===", menFiles[['2001']])

badAgeIndex = which(is.na(age2001)) + 5
menFiles[['2001']][ badAgeIndex ]
 
badAgeIndex

extractVariables = 
function(file, varNames =c("name", "home", "ag", "gun",
                           "net", "time"))
{
  
  # Find the index of the row with =s
  eqIndex = grep("^===", file)
  # Extract the two key rows and the data 
  spacerRow = file[eqIndex] 
  headerRow = tolower(file[ eqIndex - 1 ])
  body = file[ -(1 : eqIndex) ]
       # Remove footnotes and blank rows
  footnotes = grep("^[[:blank:]]*(\\*|\\#)", body)
  if ( length(footnotes) > 0 ) body = body[ -footnotes ]
  blanks = grep("^[[:blank:]]*$", body)
  if (length(blanks) > 0 ) body = body[ -blanks ]
  
  
  # Obtain the starting and ending positions of variables   
  searchLocs = findColLocs(spacerRow)
  locCols = selectCols(varNames, headerRow, searchLocs)
  
  Values = mapply(substr, list(body), start = locCols[1, ], 
                  stop = locCols[2, ])
  colnames(Values) = varNames
  
  return(Values)
}

menResMat = lapply(menFiles, extractVariables)
#womenResMat = lapply(womenFiles, extractVariables)

which(age2001 < 5)

menFiles[['2001']][ which(age2001 < 5) + 5 ]

charTime = menResMat[['2012']][, 'time']
head(charTime, 5)

tail(charTime, 5)

timePieces = strsplit(charTime, ":")

timePieces[[1]]

tail(timePieces, 1)

timePieces = sapply(timePieces, as.numeric)

runTime = sapply(timePieces, 
                 function(x) {
                   if (length(x) == 2) x[1] + x[2]/60
                   else 60*x[1] + x[2] + x[3]/60
                 })

summary(runTime)

convertTime = function(time) {
  timePieces = strsplit(time, ":")
  timePieces = sapply(timePieces, as.numeric)
  sapply(timePieces, function(x) {
                      if (length(x) == 2) x[1] + x[2]/60
                      else 60*x[1] + x[2] + x[3]/60
                      })
}

createDF = 
function(Res, year, sex) 
{
       # Determine which time to use
  useTime = if( !is.na(Res[1, 'net']) )  
              Res[ , 'net']
            else if( !is.na(Res[1, 'gun']) ) 
               Res[ , 'gun']
            else 
               Res[ , 'time']

  runTime = convertTime(useTime)
  
  Results = data.frame(year = rep(year, nrow(Res)),
                       sex = rep(sex, nrow(Res)),
                       name = Res[ , 'name'],
                       home = Res[ , 'home'],
                       age = as.numeric(Res[, 'ag']), 
                       runTime = runTime,
                       stringsAsFactors = FALSE)
  invisible(Results)
}

menDF = mapply(createDF, menResMat, year = 1999:2012,
               sex = rep("M", 14), SIMPLIFY = FALSE)

warnings()[ c(1:2, 49:50) ]

sapply(menDF, function(x) sum(is.na(x$runTime)))

createDF = function(Res, year, sex) 
{
  # Determine which time to use
  if ( !is.na(Res[1, 'net']) ) useTime = Res[ , 'net']
  else if ( !is.na(Res[1, 'gun']) ) useTime = Res[ , 'gun']
  else useTime = Res[ , 'time']
  
  # Remove # and * and blanks from time
  useTime = gsub("[#\\*[:blank:]]", "", useTime)
  runTime = convertTime(useTime[ useTime != "" ])
  
  # Drop rows with no time
  Res = Res[ useTime != "", ]
  
  Results = data.frame(year = rep(year, nrow(Res)),
                       sex = rep(sex, nrow(Res)),
                       name = Res[ , 'name'], home = Res[ , 'home'],
                       age = as.numeric(Res[, 'ag']), 
                       runTime = runTime,
                       stringsAsFactors = FALSE)
  invisible(Results)
}

menDF = mapply(createDF, menResMat, year = 1999:2012,
               sex = rep("M", 14), SIMPLIFY = FALSE)

sapply(menDF, function(x) sum(is.na(x$runTime)))

separatorIdx = grep("^===", menFiles[["2006"]])
separatorRow = menFiles[['2006']][separatorIdx]
separatorRowX = paste(substring(separatorRow, 1, 63), " ", 
                      substring(separatorRow, 65, nchar(separatorRow)), 
                      sep = "")
menFiles[['2006']][separatorIdx] = separatorRowX

menResMat = sapply(menFiles, extractVariables)
menDF = mapply(createDF, menResMat, year = 1999:2012,
               sex = rep("M", 14), SIMPLIFY = FALSE)

#separatorIdx = grep("^===", womenFiles[["2006"]])
#separatorRow = womenFiles[['2006']][separatorIdx]
#separatorRowX = paste(substring(separatorRow, 1, 63), " ", 
#                      substring(separatorRow, 65, nchar(separatorRow)), 
#                      sep = "")
#womenFiles[['2006']][separatorIdx] = separatorRowX

#womenResMat = sapply(womenFiles, extractVariables)
#womenDF = mapply(createDF, womenResMat, year = 1999:2012,
#               sex = rep("W", 14), SIMPLIFY = FALSE)

pdf("CB_BoxplotTimeByYr.pdf", width = 8, height = 5)
boxplot(sapply(menDF, function(x) x$runTime), 
        xlab = "Year", ylab = "Run Time (min)")
dev.off()

cbMen = do.call(rbind, menDF)
save(cbMen, file = "cbMen.rda")

dim(cbMen)

load("cbMen.rda")

pdf("CB_Overplot.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

plot(runTime ~ age, data = cbMen, ylim = c(40, 180),
     xlab = "Age (years)", ylab = "Run Time (minutes)")

par(oldPar)
dev.off()

library(RColorBrewer)
ls("package:RColorBrewer")

display.brewer.all()

Purples8 = brewer.pal(9, "Purples")[8]
Purples8

Purples8A = paste(Purples8, "14", sep = "")

pdf("CB_OverplotTransparent.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))
plot(runTime ~ jitter(age, amount = 0.5), 
     data = cbMen, 
     pch = 19,cex = 0.2, col = Purples8A,
     ylim = c(45, 165), xlim = c(15, 85),
     xlab = "Age (years)", ylab = "Run Time (minutes)")
par(oldPar)
dev.off()

pdf("CB_SmoothScatter.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

smoothScatter(y = cbMen$runTime, x = cbMen$age,
              ylim = c(40, 165), xlim = c(15, 85),
              xlab = "Age (years)", ylab = "Run Time (minutes)")

par(oldPar)
dev.off()

cbMenSub = cbMen[cbMen$runTime > 30 &
                 !is.na(cbMen$age) & cbMen$age > 15, ]

ageCat = cut(cbMenSub$age, breaks = c(seq(15, 75, 10), 90))
table(ageCat)

pdf("CB_Boxplots.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

plot(cbMenSub$runTime ~ ageCat, 
     xlab = "Age (years)", ylab = "Run Time (minutes)")

par(oldPar)
dev.off()

lmAge = lm(runTime ~ age, data = cbMenSub)

lmAge$coefficients

summary(lmAge)

class(lmAge)

pdf("CB_ResidSimpleLM.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

smoothScatter(x = cbMenSub$age, y = lmAge$residuals,
              xlab = "Age (years)", ylab = "Residuals")
abline(h = 0, col = "purple", lwd = 3)
 
resid.lo = loess(resids ~ age, 
                 data = data.frame(resids = residuals(lmAge),
                                   age = cbMenSub$age))

age20to80 = 20:80

resid.lo.pr = 
  predict(resid.lo, newdata = data.frame(age = age20to80))

lines(x = age20to80, y = resid.lo.pr, col = "green", lwd = 2)
par(oldPar)
dev.off()

menRes.lo = loess(runTime ~ age, cbMenSub)

menRes.lo.pr = predict(menRes.lo, data.frame(age = age20to80))

over50 = pmax(0, cbMenSub$age - 50)

lmOver50 = lm(runTime ~ age + over50, data = cbMenSub)

summary(lmOver50)

decades = seq(30, 60, by = 10)
overAge = lapply(decades, 
                 function(x) pmax(0, (cbMenSub$age - x)))
names(overAge) = paste("over", decades, sep = "")
overAge = as.data.frame(overAge)
tail(overAge)

lmPiecewise = lm(runTime ~ . , 
                 data = cbind(cbMenSub[, c("runTime", "age")], 
                              overAge))

summary(lmPiecewise)

overAge20 = lapply(decades, function(x) pmax(0, (age20to80 - x)))
names(overAge20) = paste("over", decades, sep = "")
overAgeDF = cbind(age = data.frame(age = age20to80), overAge20)

tail(overAgeDF)

predPiecewise = predict(lmPiecewise, overAgeDF)

plot(predPiecewise ~ age20to80,
     type = "l", col = "purple", lwd = 3,
     xlab = "Age (years)", ylab = "Run Time Prediction")

lines(x = age20to80, y = menRes.lo.pr, 
      col = "green", lty = 2, lwd = 3)
legend("topleft", col = c("purple", "green"),
       lty = c(1, 2), lwd= 3,
       legend = c("Piecewise Linear", "Loess Curve"), bty = "n")

pdf("CB_PiecewiseLoessCurves.pdf", width = 8, height = 6)
plot(predPiecewise ~ age20to80,
     type = "l", col = "#984ea3", lwd = 3,
  #   type = "l", col = "purple", lwd = 2,
     xlab = "Age (years)", ylab = "Run Time Prediction")

lines(x = age20to80, y = menRes.lo.pr, col = "#4daf4a", lwd = 3, lty = 2)
legend("topleft", col = c("#984ea3", "#4daf4a"), lty = c(1, 2), lwd = 3,
       legend = c("Piecewise Linear", "Loess Curve"), bty = "n")

#lines(x = age20to80, y = menRes.lo.pr, col = "green", lwd = 2)
#legend("topleft", col = c("purple", "green"), lty = 1, lwd = 2,
#       legend = c("Piecewise Linear", "Loess Curve"), bty = "n")
dev.off()


pdf("CB_NumRunnersLinePlot.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

numRunners = with(cbMen, tapply(runTime, year, length))
plot(numRunners ~ names(numRunners), type="l", lwd = 2,
     xlab = "Years", ylab = "Number of Runners")
par(oldPar)
dev.off()

summary(cbMenSub$runTime[cbMenSub$year == 1999])

summary(cbMenSub$runTime[cbMenSub$year == 2012])

pdf("CB_AgeDensity99vs12.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

age1999 = cbMenSub[ cbMenSub$year == 1999, "age" ]
age2012 = cbMenSub[ cbMenSub$year == 2012, "age" ]

plot(density(age1999, na.rm = TRUE), 
     ylim = c(0, 0.05), col = "purple",
     lwd = 3,  xlab = "Age (years)",  main = "")
lines(density(age2012, na.rm = TRUE), 
      lwd = 3, lty = 2, col="green")
legend("topleft", col = c("purple", "green"), lty= 1:2, lwd = 3,
       legend = c("1999", "2012"), bty = "n")

par(oldPar)
dev.off()

qqplot(age1999, age2012, pch = 19, cex = 0.5, 
       ylim = c(10,90), xlim = c(10,90), 
       xlab = "Age in 1999 Race",
       ylab = "Age in 2012 Race", 
       main = "Quantile-quantile plot of male runner's age")
abline(a =0, b = 1, col="red", lwd = 2)

mR.lo99 = loess(runTime ~ age, cbMenSub[ cbMenSub$year == 1999,])
mR.lo.pr99 = predict(mR.lo99, data.frame(age = age20to80))

mR.lo12 = loess(runTime ~ age, cbMenSub[ cbMenSub$year == 2012,])
mR.lo.pr12 = predict(mR.lo12, data.frame(age = age20to80))

plot(mR.lo.pr99 ~ age20to80,
     type = "l", col = "purple", lwd = 3,
     xlab = "Age (years)", ylab = "Fitted Run Time (minutes)")
   
lines(x = age20to80, y = mR.lo.pr12,
      col = "green", lty = 2, lwd = 3)
 
legend("topleft", col = c("purple", "green"), lty = 1:2, lwd = 3,
       legend = c("1999", "2012"), bty = "n")


pdf("CB_Loess99vs12.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

plot(mR.lo.pr99 ~ age20to80,
     type = "l", col = "#984ea3", lwd = 3,
     xlab = "Age (years)", ylab = "Prediction (minutes)")  
lines(x = age20to80, y = mR.lo.pr12, col="#4daf4a", lty = 2, lwd = 3) 
legend("topleft", col = c("#984ea3", "#4daf4a"), lty = 1:2, lwd = 3,
       legend = c("1999", "2012"), bty = "n")
par(oldPar)
dev.off()

gap14 = mR.lo.pr12 - mR.lo.pr99

pdf("CB_DifferenceInFittedCurves.pdf", width = 8, height = 6)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

plot(gap14 ~ age20to80, type = "l" , xlab = "Age (years)", 
     ylab = "Difference in Fitted Curves (minutes)", lwd = 2)
par(oldPar)
dev.off()

fastestMan = tapply(menRes$time, menRes$age, min, na.rm = TRUE)
plot(fastestMan ~ names(fastestMan), type ="l", xlim = c(20, 80))
ageFM = as.numeric(names(fastestMan))
mR.loF = loess(fastestMan ~ ageFM)
mR.lo.prF = predict(mR.loF, data.frame(age = ageFM), se = FALSE)
lines(x = ageFM, y = mR.lo.prF, col = "purple", lwd = 2)

timeNorm = menRes$time / mR.lo.prF[as.character(menRes$age)]
time99Norm = timeNorm[menRes$year == 1999]
time12Norm = timeNorm[menRes$year == 2012]
summary(time99Norm)

summary(time12Norm)

plot(density(100*time99Norm, na.rm = TRUE), 
    # ylim = c(0, 0.05), 
     col = "purple",
     lwd = 3,  xlab = "Time (percentage)",
     main = "Time Distribution for 1999 and 2012 Runners\n Percentage of the fastest runner for that age")
lines(density(100*time12Norm, na.rm = TRUE), 
          lwd = 3, col = "green")
legend("topleft", fill = c("purple", "green"),
       legend = c("1999", "2012"), bty = "n")


trimBlanks = function(charVector) {
  nameClean = gsub("^[[:blank:]]+", "", charVector)
  nameClean = gsub("[[:blank:]]+$", "", nameClean)
  nameClean = gsub("[[:blank:]]+", " ", nameClean)
}

nameClean = trimBlanks(cbMenSub$name)

length(nameClean)

length(unique(nameClean))

table(table(nameClean))

head( sort(table(nameClean), decreasing = TRUE), 1)

mSmith = cbMenSub[nameClean == "Michael Smith", ]

head(unique(mSmith$home))

nameClean = tolower(nameClean)

head( sort(table(nameClean), decreasing = TRUE), 1)

nameClean = gsub("[,.]", "", nameClean)

tabNameYr = table(cbMenSub$year, nameClean)

max(tabNameYr)

class(tabNameYr)

mode(tabNameYr)

names(attributes(tabNameYr))

dim(tabNameYr)

head(colnames(tabNameYr), 3)

which( tabNameYr == max(tabNameYr) )

which( tabNameYr == max(tabNameYr), arr.ind = TRUE )

indMax = which( tabNameYr == max(tabNameYr), arr.ind = TRUE )
colnames(tabNameYr)[indMax[2]]

cbMenSub$nameClean = nameClean

cbMenSub$yob = cbMenSub$year - cbMenSub$age

# Fix home in a similar way
homeClean = trimBlanks(tolower(cbMenSub$home))
cbMenSub$homeClean = gsub("[,.]", "", homeClean)

vars = c("year", "homeClean", "nameClean", "yob",  "runTime")
mb = which(nameClean == "michael brown")
birthOrder = order(cbMenSub$yob[mb])
cbMenSub[mb[birthOrder], vars]

cbMenSub$ID = paste(nameClean, cbMenSub$yob, sep = "_")

races = tapply(cbMenSub$year, cbMenSub$ID, length)

races8 = names(races)[which(races >= 8)]

men8 = cbMenSub[ cbMenSub$ID %in% races8, ]

orderByRunner = order(men8$ID, men8$year)
men8 = men8[orderByRunner, ]

men8L = split(men8, men8$ID)
names(men8L) = races8

length(unique(men8$ID))

gapTime = tapply(men8$runTime, men8$ID,
                 function(t) any(abs(diff(t)) > 20))

gapTime = sapply(men8L, function(df) 
                          any(abs(diff(df$runTime)) > 20))

sum(gapTime)

lapply(men8L[ gapTime ][1:2], function(df) df[, vars])

homeLen = nchar(cbMenSub$homeClean)

cbMenSub$state = substr(cbMenSub$homeClean, 
                        start = homeLen - 1, stop = homeLen)

cbMenSub$state[cbMenSub$year == 2006] = NA

cbMenSub$ID = paste(cbMenSub$nameClean, cbMenSub$yob, 
                    cbMenSub$state, sep = "_")

numRaces = tapply(cbMenSub$year, cbMenSub$ID, length)
races8 = names(numRaces)[which(numRaces >= 8)]
men8 = cbMenSub[ cbMenSub$ID %in% races8, ]
orderByRunner = order(men8$ID, men8$year)
men8 = men8[orderByRunner, ]

men8L = split(men8, men8$ID)
names(men8L) = races8

length(races8)

groups = 1 + (1:length(men8L) %% 9)

addRunners = function(listRunners, colors, numLty) 
{
  numRunners = length(listRunners)
  colIndx = 1 + (1:numRunners) %% length(colors)
  ltys = rep(1:numLty, each = length(colors), length = numRunners)

  mapply(function(df, i) {      
           lines(df$runTime ~ df$age, 
           col = colors[colIndx[i]], lwd = 2, lty = ltys[i])
         }, listRunners, i = 1:numRunners) 
}

colors = c("#e41a1c", "#377eb8","#4daf4a", "#984ea3", 
           "#ff7f00", "#a65628")
par(mfrow = c(3, 3), mar = c(2, 2, 1, 1))
invisible(
  sapply(1:9, function(grpId){
    plot( x = 0, y = 0, type = "n",
          xlim = c(20, 80), ylim = c(50, 130),
          xlab = "Age (years)", ylab = "Run Time (minutes)")
     
    addRunners(men8L[ groups == grpId ], colors, numLty = 6)
  }) )

fitOne = function(oneRunner, addLine = FALSE, col = "grey") {
  lmOne = lm(runTime ~ age, data = oneRunner)
  if (addLine) 
    lines(x = oneRunner$age, y = predict(lmOne), 
          col = col, lwd = 2, lty = 2)

  ind = floor( (nrow(oneRunner) + 1) / 2)
  res = c(coefficients(lmOne)[2], oneRunner$age[ind],
          predict(lmOne)[ind])
  names(res) = c("ageCoeff", "medAge", "predRunTime")
  return(res)
}

par(mfrow = c(1, 1), mar = c(5, 4, 1, 1))

plot( x = 0, y = 0, type = "n",
      xlim = c(20, 80), ylim = c(50, 130),
      xlab = "Age (years)", ylab = "Run Time (minutes)")
     
addRunners(men8L[ groups == 9 ], colors, numLty = 6)
lapply(men8L[groups == 9], fitOne, addLine = TRUE, col = "black")

men8LongFit = lapply(men8L, fitOne)

coeffs = sapply(men8LongFit, "[", "ageCoeff" )
ages = sapply(men8LongFit, "[", "medAge")

longCoeffs = lm(coeffs ~ ages)

summary(longCoeffs)

pdf("CB_LongCoeffs.pdf", width = 10, height = 7)
oldPar = par(mar = c(4.1, 4.1, 1, 1))
plot(coeffs ~ ages, xlab = "Median Age (years)",
     ylab = "Coefficient (minutes per race / year)")
abline(longCoeffs, col = "#984ea3", lwd = 3)
abline(h = 0, col="blue", lwd = 3)
loCoeffs = loess(coeffs ~ ages)
ageV = min(ages):max(ages)
predV = predict(loCoeffs, new = data.frame(ages = ageV))
lines(x = ageV, y = predV, lwd = 3, lty = 2, col = "#4daf4a")
par(oldPar)
dev.off()

